---
title: "Statistical Models for Time Series"
author: "Della"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statistical Models for Time Series

We will map some linear statistical models to **time series**. These models are related to linear regression, but represent correlations between data points in the same **time series**, 
unlike standard methods applied to cross-sectional data, where each data point is assumed to be independent of the others in the sample. We will analyze the following models:
  - autoregressive models (AR), moving average models (MA) and models 
  autoregressive integrated moving average (ARIMA);
  - vector autoregression (VAR);
  - hierarchical models;
Traditionally, these models have been the driving force in forecasting **time series** and continue to be used in a wide range of situations, from academic research to modeling in various fields of acting


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
